On the propriety of restricted Boltzmann machines

### 2016-11-01 writing/draft.Rmd
Dr. Vardeman's edits

---
title: "On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "September, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has been largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability. We also describe the difficulties that arise in likelihood-based fitting and Bayes of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
---

@@

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable. The literature often suggests Markov chain Monte Carlo (MCMC) approximations to posteriors to fit these models (e.g., Gibbs sampling to exploit conditional structure in the hidden and visible variables), but little is said about acheiving convergence [@hinton2010practical, @hinton2006fast]. Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) per observation (visibles) which can critically influence convergence in MCMC-based likelihood methods.

In Section \ref{bayesian-model-fitting}, we compare three Bayesian modeling techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.

Additionally, because RBM models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the (discrete) empirical distribution of most realistic training sets. That is, there is no "smoothed distribution" achieved in fitting a RBM model with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows. Section \ref{background} formally definesthe RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of detecting it are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques intended to avoid fitted model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}. Additionally, there is an accompanying appendix with proofs of main results.

@@

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in an RBM for some integers $\nv, \nh \ge 1$. Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$ with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural or na{\"i}ve starting point, we show later that the coding $\mathcal{C}=\{-1,1\}$ induces better model properties for the RBM. We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes as

\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align}

@@

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion. A kind of Gibbs sampling can be tried, which is natural due to conditional independence architecture of the RBM (i.e. visibles given hiddens or vice versa). Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer allows for stepwise simulation of both the visible and hidden layers.

@@

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Again, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t(X)$, is close to the boundary of $C$.

The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows. When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\}.
$$

@@

Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*. Instability can be roughly characterized as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how the RBM model might be expanded to incorporate more and more visibles When increasing the size of a RBM model, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{V} \equiv (\theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}, \theta_{11}, \dots, \theta_{VH}), V \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
\begin{align}
\frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}\right] \equiv \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_V}(v_1, \dots, v_V) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^V \theta_{v_i} v_i + \sum_{j = 1}^H \theta_{h_j} h_j + \sum_{i = 1}^V \sum_{j = 1}^H \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_V)$ for the visible variables $(V_1, \dots, V_V)$ under parameters $\boldsymbol \theta_V$.

In formulating a RBM model for a potentially large number of visibles (i.e., $V \rightarrow \infty$), we will say that the ratio (\ref{eqn:elpr}) needs to stay bounded for the sequence of RBM models to be stable. That is, we make the following convention

\begin{defn}[S-unstable RBM]
Let $\boldsymbol \theta_V \in \mathbb{R}^{V + H +  H*V}, v \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $H \equiv H(V) \ge 1$, can be an arbitrary function of the number $V$ of visibles. A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{V \rightarrow \infty} \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \lim\limits_{V \rightarrow \infty} \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $V_C > 0$ so that $\frac{1}{V}\text{ELPR}(\boldsymbol \theta_V) > C$ for all $V \ge V_C$.

This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_V)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_V$ is not fixed, but rather grows with $V$.

S-unstable RBM model sequences are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_V) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*)} : (v_1, \dots, v_V) \& (v_1^*, \dots, v_V^*) \in \mathcal{C}^V \text{ differ by exactly one component}\right\},
$$
denoting the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_V$. We then have the following result.

\begin{prop}
\label{prop:instab}
Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_V)$ be as in (\ref{eqn:elpr}) for an integer $V \ge 1$. If $\frac{1}{V}\text{ELPR}_V(\boldsymbol \theta) > C$, then $\Delta_V(\boldsymbol \theta) > C$.
\end{prop}

@@

Additionally, S-unstable RBM model sequences are connected to degenerate model sequences placing all probability on a small portion of their sample spaces. To see this, define a modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_V, V \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_V} \equiv \left\{(v_1, \dots, v_V) \in \mathcal{C}^V: \log P_{\boldsymbol \theta_V}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then S-unstable model sequences are guaranteed to be degenerate, as the following result shows.

\begin{prop}
\label{prop:degen}
For an S-unstable RBM model and any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_V}\left((v_1, \dots, v_V) \in M_{\epsilon, \boldsymbol \theta_V}\right) \rightarrow 1$ holds as $V \rightarrow \infty$.
\end{prop}

@@

\begin{remark}
\label{remark:bnd}
To later numerically investigate S-unstable RBM model sequences, it is useful to define a computationally tractable lower bound on (\ref{eqn:elpr}) given by
\begin{align}
\frac{1}{V}R(\boldsymbol \theta_V) \text{ for } R(\boldsymbol \theta_V) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q_{\boldsymbol \theta_V}(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q_{\boldsymbol \theta_V}(\boldsymbol x) - H\log 2. \label{eqn:r}
\end{align}
That is, in (\ref{eqn:elpr}),
\begin{align}
\frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \ge \frac{1}{V}R(\boldsymbol \theta_V). \label{eqn:lowerbnd}
\end{align}
\end{remark}
The appendix provides a proof of (\ref{eqn:lowerbnd}). The benefit of $R(\boldsymbol \theta_V)$ is that $R(\boldsymbol \theta_V)/V$ is simple to compute (i.e. unlike (\ref{eqn:elpr}), requiring no summation over hidden variables) and that Propositions \ref{prop:instab}-\ref{prop:degen} continue to hold upon replacing $\text{ELPR}(\boldsymbol \theta_V)/V$ with $R(\boldsymbol \theta_V)/V$ due to (\ref{eqn:lowerbnd}). Hence, large values of $R(\boldsymbol \theta_V)/V$ also indicate S-unstable RBM model sequences.

>>>>

---
title: "On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "September, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has been largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability. We also describe the difficulties that arise in likelihood-based and Bayes fitting  of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
---

@@

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable. The literature often suggests Markov chain Monte Carlo (MCMC) approximations to posteriors to fit these models (e.g., Gibbs sampling to exploit conditional structure in the hidden and visible variables), but little is said about acheiving convergence [@hinton2010practical, @hinton2006fast]. Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.

In Section \ref{bayesian-model-fitting}, we compare three Bayesian modeling techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield unattractive models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.

Additionally, because RBM models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the (discrete) empirical distribution of most realistic training sets. That is, there is no "smoothed distribution" achieved in fitting a RBM model with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way.

This paper is structured as follows. Section \ref{background} formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of detecting it are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques intended to avoid fitted model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}. Additionally, there is an accompanying appendix with proofs of main results.

@@

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in an RBM for some integers $\nv, \nh \ge 1$. Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$ with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural or naive starting point, we show later that the coding $\mathcal{C}=\{-1,1\}$ induces better model properties for the RBM. We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes as

\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align}

@@

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic. A kind of Gibbs sampling can be tried (which is natural due to conditional independence architecture of the RBM, i.e. visibles given hiddens or vice versa). Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer allows for stepwise simulation of both hidden layers and model parameters possible.

@@

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy*, *instability*, and *uninterpretability*. In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.

The RBM class of models can be thought to exhibit an analogous form of *near-degeneracy* as follows. When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
$$

@@

Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*. Instability can be roughly characterized as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how the RBM model might be expanded to incorporate more and more visibles. When increasing the size of a RBM model, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameters $\boldsymbol \theta_\nv$.

In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\ref{eqn:elpr}) needs to stay bounded for the sequence of RBM models to be stable. That is, we make the following convention

\begin{defn}[S-unstable RBM]
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \equiv \lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta}(v_1, \dots, v_\nv)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $n_C > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$ for all $\nv \ge n_C$.

This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.

S-unstable RBM model sequences are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{P_{\boldsymbol \theta_\nv}(v_1^*, \dots, v_\nv^*)} : (v_1, \dots, v_\nv) \& (v_1^*, \dots, v_\nv^*) \in \mathcal{C}^\nv \text{ differ by exactly one component}\right\},
$$
denoting the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.

\begin{prop}
\label{prop:instab}
Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\ref{eqn:elpr}) for an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$, then $\Delta(\boldsymbol \theta_\nv) > C$.
\end{prop}

@@

Additionally, S-unstable RBM model sequences are connected to degenerate model sequences placing all probability on a small portion of their sample spaces. To see this, define a modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_\nv, \nv \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then S-unstable model sequences are guaranteed to be degenerate, as the following result shows.

\begin{prop}
\label{prop:degen}
For an S-unstable RBM model sequence and any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1$ as $\nv \rightarrow \infty$.
\end{prop}

@@

### Uninterpretability
For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by the model statement. We adapt this notion to RBM models as follows. Note that, in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nature of dependence) can break down as interaction parameters increase in magnitude relative to the size of the main effects. In such case, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the modes (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (destroying the model interpretation of dependence as local conditional modifications of all overall marginal mean structure). To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters. Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence. Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models. If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable. The quantities to compare in the RBM case are
